Local Type Inference for Polarised System F with Existentials

### 2024-05-07 oopsla24/tex/04-algorithm.tex
Edit: sections 1-5

We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish them from
normal variables $[[α⁺]]$, $[[β⁻]]$.
In a few places, we replace the quantified variables $[[pas]]$ 
with their algorithmic counterpart $[[puas]]$.
The procedure of replacing declarative variables with algorithmic ones we call 
\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$.

@@

@@

\paragraph{Auxiliary Functions}
We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of algorithmic variables
that it restricts. Similarly, we define $[[dom(Θ)]]$---the domain of instantiation context
---as a set of algorithmic variables that $[[Θ]]$ associates with their contexts.
We write $[[Θ(α̂±)]]$ to denote the declarative context associated with $[[α̂±]]$ in $[[Θ]]$.

@@

  Some of the procedures (in particular, the unification and the
  anti-unification) assume that the input types are normalized. Therefore, we
  call the normalization procedure before invoking them, indicating this by the
  `$\ottkw{nf}$' annotation on the arrows in \cref{fig:alg-subtyping-graph}.
  Alternatively, one could normalize the input types in the very beginning,
  as the normalization is preserved throughout the algorithm. However, we
  delay the normalization to the places where it is required to show that
  normalization is needed only at these stages to maintain consistent
  invariants.

@@

In general, we must 
\begin{itemize*}
  \item[(i)] find the most detailed pattern (a type with `holes' at negative positions) 
    that matches both sides, and 
  \item[(ii)] abstract over the `holes' by existential quantifiers.
\end{itemize*}
The algorithm that finds the most detailed common pattern is called \emph{anti-unification}.
As output, it returns $[[(Ξ, uP, aus1, aus2)]]$, where important for us is
$[[uP]]$---the pattern and $[[Ξ]]$---the set of `holes' represented by negative algorithmic variables.
We discuss the anti-unification algorithm in detail in the following section.

@@

The anti-unification algorithm \cite{plotkin1970:generalization,reynolds1970:transform}, 
is a procedure dual to unification
For two given (potentially different) expressions, 
it finds the most specific generalizer---the most detailed pattern that 
matches both of the input expressions. As evidence, it can also return 
two substitutions that instantiate the `holes' of the pattern to the input expressions.

In our case, we have to be more demanding on the anti-unification algorithm.
Since we use it to construct an existential type, whose (negative) quantified variables
can only be instantiated with negative types, we must make sure that the pattern 
has `holes' only at negative positions. Moreover, we must make sure that
the resulting substitutions for the `holes' are well-formed in the initial contest, 
and do not contain variables bound later. For example, 
the anti-unification of $[[iN1 = ∀β⁺.α1⁺ → ↑β⁺]]$ and
$[[iN2 = ∀β⁺.α2⁺ → ↑β⁺]]$
must result in a `hole', which we model as an algorithmic type variable $[[γ̂⁻]]$, 
with a pair of substitutions $[[γ̂⁻ ↦ iN1]]$ and $[[γ̂⁻ ↦ iN2]]$.
But it cannot be more specific such as
$[[∀β⁺.γ̂⁺ → ↑β⁺]]$ (since the hole cannot be positive)
or $[[∀β⁺.γ̂⁻]]$ 
(since the instantiation cannot capture the bound variable $[[β⁺]]$).

@@

  Second, this rule is only applicable to negative types, 
  moreover, the input types are checked to be well-formed 
  in the outer context $[[Γ]]$. This is required by the 
  usage of anti-unification: we 
  call it to build an existential type that would be 
  an upper bound of two input types via abstracting
  some of their subexpressions under existential quantifiers.
  The existentials quantify over \emph{negative} variables,
  and they must be instantiated in the context
  available at that moment.

  Third, the rule is only applicable if all other rules fail. 
  Notice that it could happen even when the input types have matching constructors. 
  For example, the generalizer of $[[↑α⁺]]$ and
  $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$),
  rather than $[[↑γ̂⁺]]$. This way, the algorithm must try to 
  apply the congruent rules first, and only if they fail,
  apply \ruleref{\ottdruleAUAULabel}.
  This principle makes the inference system not syntax-directed:
  it is not known a priori (before the recursive call) 
  whether the corresponding congruent rule or rule \ruleref{\ottdruleAUAULabel} 
  will be applied.

@@

This way, the inference algorithm also consists of three mutually recursive procedures: 
the positive type inference, the negative type inference, and the
application type inference. As subroutines, the inference algorithm uses
subtyping, constraint merge, and normalization.  
The corresponding graph is shown in \cref{fig:alg-typing-graph}.

\begin{figure}[h]
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.2cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
    % Define nodes
    \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
    \node[right=1.2cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
    \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
    \node[below=0.6cm of 2] (7) {Minimal Instantiation\\$[[Γ ⊢ uP SC minby uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[below=0.5cm of 7] (8) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \node[below=0.9cm of 6] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
    \draw[->] ([yshift=+.2cm]1.east) to ([yshift=+.2cm]2.west);
    \draw[->, bend right=20] (3) to (1);
    \draw[->] (2) to (3);
    \draw[->] (2) to (4);
    \draw[->] (2) to (7);
    \draw[->] (3) to (4);
    \draw[->] (7) to (8);
    \draw[->] (1.south west) |- (5);
    \draw[->] (3) |- (5.south east);
    \draw[->] (2) to (6);
  \end{tikzpicture}  
  \caption{Dependency graph of the typing algorithm}
  \label{fig:alg-typing-graph}
\end{figure}

@@

@@

  To ensure that this step does not cause infinite recursion, 
  we also check that the head type has at least one 
  $[[∀]]$-quantifier. Also, to force the algorithm to
  apply rule \ruleref{\ottdruleATEmptyAppLabel} when  
  there are no arguments, we require $[[args ≠ ·]]$.

\paragraph{Application of an Arrow Type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the minimal constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.

@@

Multiple types $[[iM]]$ can be inferred for a type application:
$[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ but only one \emph{principal} 
type should be chosen for a variable in an unannotated let binder.
Declaratively, we require the principal type $[[iP]]$ to be
minimal among other all types $[[iP']]$ that can be inferred
for the application $[[Γ ; Φ ⊢ iN ● args ⇒> ↑iP']]$.
Algorithmically, the inference returns an
\emph{algorithmic} type $[[uP]]$ 
together with a set of necessary and sufficient constraints $[[SC]]$
that any instantiation of $[[uP]]$ must satisfy.
This way, to guarantee principality, we must minimize the instantiation of $[[uP]]$,
while ensuring that it satisfies the given constraints $[[SC]]$.

The minimal instantiation judgment `$[[Γ ⊢ uP SC minby uσ]]$'
means that $[[uσ]]$ is the minimum among all the substitutions satisfying $[[SC]]$,
that is $[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
instantiation of $[[uP]]$. The algorithm is given in \cref{fig:minimal-instantiation}.
First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
Then it considers two cases. If the type is an algorithmic variable $[[α̂⁺]]$
restricted by a constraint $[[(α̂⁺ :≥ iP) ∊ SC]]$,
its minimal instantiation is the type $[[nf(iP)]]$, 
and \ruleref{\ottdruleSINGPUvarLabel} is applied.
Otherwise, \ruleref{\ottdruleSINGSingLabel} is applied,
meaning that all the instantiations of the type must be equivalent.
The algorithm ensures that by two premises:
\begin{enumerate*}
  \item[(i)] checking that all the algorithmic variables of the type are restricted by the constraint set; and
  \item[(ii)] building the substitution that satisfies the constraint set on these variables 
    (and simultaneously checking that such substitution is unique up-to-equivalence)
    using the constraint singularity algorithm.
\end{enumerate*}

@@

To implement the singularity algorithm, we define a partial function $[[SC
singular with uσ]]$, taking a subtyping constraint $[[SC]]$ as an argument and
returning a substitution $[[uσ]]$---the only solution of $[[SC]]$. The
constraint $[[SC]]$ consists of constraint entries. Thus, we define the
singularity using the singularity of constraint entries: each entry of $[[SC]]$
must have a unique instantiation and the resulting substitution $[[uσ]]$ must be
the combination of these instantiations (see \cref{fig:constraint-singularity}).

\begin{figure}[h]
  For a constraint set $[[SC]]$, we write $[[SC singular with uσ]]$ if and only if:
  \begin{itemize}
    \item for any \emph{positive} constraint entry $[[scE ∊ SC]]$, there exists $[[iP]]$ such that $[[scE singular with iP]]$,
      and for the variable $[[β̂⁺]]$ restricted by $[[scE]]$, $[[ [uσ]β̂⁺ = iP ]]$;
    \item the symmetric property holds for all \emph{negative} $[[scE ∊ SC]]$; and
    \item for any $[[α̂± ∉ dom(SC)]]$, $[[ [uσ]α̂± = α̂± ]]$.
  \end{itemize}
  \caption{Constraint Singularity}
  \label{fig:constraint-singularity}
\end{figure}

The singularity of constraint \emph{entries} is defined in
\cref{fig:constraint-entry-singularity}. The \emph{equivalence} entries are
always singular, as the only possible type satisfying them is the one given in
the constraint itself, which is reflected in \ruleref{\ottdruleSINGNEqLabel} and
\ruleref{\ottdruleSINGPEqLabel}. The \emph{subtyping} constraints are trickier.
As will be discussed in \cref{sec:proof-lub-upgrade}, variables 
(and equivalent types such as $[[∃nas.pa]]$) do not have 
\emph{proper} supertypes, and thus, the constraints of a form 
$[[pua :≥ ∃nas.pa]]$ are singular with the only possible normalized solution $[[pa]]$
\ruleref{\ottdruleSINGSupVarLabel}. However, if the body of the existential type
is guarded by a shift $[[∃nas.↓iN]]$, it is singular if and only if $[[iN]]$ is
equivalent to some $[[αi⁻ ∊ {nas}]]$ bound by the quantifier, see the rule
\ruleref{\ottdruleSINGSupShiftLabel}.

\begin{figure}[h]
  \hfill\\
  \ottdefnSINGLabeled
  \caption{Singular Constraint Entry}
  \label{fig:constraint-entry-singularity}
\end{figure}

>>>>

We denote algorithmic variables as $[[α̂⁺]]$, $[[β̂⁻]]$, \dots to distinguish
them from normal variables $[[α⁺]]$, $[[β⁻]]$. In a few places, we replace the
quantified variables $[[pas]]$ with their algorithmic counterpart $[[puas]]$.
The procedure of replacing declarative variables with algorithmic ones we call
\emph{algorithmization} and denote as $[[ nuas/nas ]]$ and $[[ puas/pas ]]$. The
converse operation---\emph{dealgorithmization}---is denoted as $[[ nas/nuas ]]$ and 
is used in the least upper bound procedure (\cref{sec:lub}).

@@

\paragraph{Algorithmic Substitution}
We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
variables to \emph{declarative} types. The signature $[[Θ ⊢ uσ : Ξ]]$
specifies the domain and the range of $[[uσ]]$: for each variable $[[α̂±]]$
in $[[Ξ]]$, there exists an corresponding entry in $[[Θ]]$ associating 
$[[α̂±]]$ with a declarative context $[[Γ]]$ such that $[[ [uσ]α̂± ]]$
is well-formed in $[[Γ]]$. In addition, we assume that $[[uσ]]$
acts as the identity on the variables not in $[[Ξ]]$.

@@

\paragraph{Auxiliary Functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv(uN)]]$ and $[[uv(uP)]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

\paragraph{Constraint well-formedness and satisfaction}
Suppose that $[[Θ]]$ is an instantiation context. 
We say that constraint set $[[SC]]$ is well-formed in $[[Θ]]$ 
(denoted as $[[Θ ⊢ SC]]$) 
if for every every constrain entry $[[scE ∊ SC]]$
associating a variable $[[α̂±]]$ with a type $[[iP]]$,
this type is well-formed in the corresponding declarative context $[[Θ(α̂±)]]$.

Substitution $[[uσ]]$ satisfies a constraint \emph{entry} $[[scE]]$ restricting
algorithmic variable $[[α̂±]]$ if
$[[ [σ]α̂± ]]$ can be validly substituted for $[[α̂±]]$ in $[[scE]]$ 
(so that the corresponding equivalence or subtyping holds).

Substitution $[[uσ]]$ satisfies a constraint \emph{set}
if $[[Θ ⊢ SC]]$ and \emph{each entry} of $[[SC]]$ is satisfied by $[[uσ]]$.

@@

  Some of the procedures (in particular, the unification and the
  anti-unification) assume that the input types are normalized. Therefore, we
  call the normalization procedure before invoking them, indicating this by the
  `$\ottkw{nf}$' annotation on the arrows in \cref{fig:alg-subtyping-graph}.
  Alternatively, one could normalize the input types in the very beginning
  and preserve the normalization throughout the algorithm. However, we
  delay the normalization to the places where it is required to show that
  normalization is needed only at these stages to maintain consistent
  invariants.

@@

In general, we must 
\begin{itemize*}
  \item[(i)] find the most detailed pattern (a type with `holes' at negative positions) 
    that matches both sides, and 
  \item[(ii)] abstract over the `holes' by existential quantifiers.
\end{itemize*}
The algorithm that finds the most detailed common pattern is called \emph{anti-unification}.
As output, it returns $[[(Ξ, uP, aus1, aus2)]]$, where important for us is
$[[uP]]$---the pattern---and $[[Ξ]]$--the set of `holes' represented by negative algorithmic variables.
We discuss the anti-unification algorithm in detail in the following section.

@@

The anti-unification algorithm
\cite{plotkin1970:generalization,reynolds1970:transform}, is a procedure dual to
unification. For two given (potentially different) expressions, it finds the most
specific generalizer---the most detailed pattern that matches both of the input
expressions. As evidence, it can also return two substitutions that instantiate
the `holes' of the pattern to the input expressions.

In our case, we have to be more demanding on the anti-unification algorithm.
Since we use it to construct an existential type, whose (negative) quantified
variables can only be instantiated with negative types, we must make sure that
the pattern has `holes' only at negative positions. Moreover, we must make sure
that the resulting substitutions for the `holes' are well-formed in the context
from the past---at the moment when the corresponding polymorphic variables were
introduced---and do not contain variables bound later. For example, the
anti-unification of $[[iN1 = ∀β⁺.α1⁺ → ↑β⁺]]$ and $[[iN2 = ∀β⁺.α2⁺ → ↑β⁺]]$ is a
singleton `hole', which we model as an algorithmic type variable $[[γ̂⁻]]$, with
a pair of substitutions $[[γ̂⁻ ↦ iN1]]$ and $[[γ̂⁻ ↦ iN2]]$. But it
\emph{cannot} be more specific such as $[[∀β⁺.γ̂⁺ → ↑β⁺]]$ (since the hole
cannot be positive) or $[[∀β⁺.γ̂⁻]]$ (since \emph{the instantiation of $~[[γ̂⁻]]$ cannot capture the
bound variable $[[β⁺]]$}).

@@

  Second, this rule is only applicable to negative types; moreover, the input
  types are checked to be well-formed in the outer context $[[Γ]]$. This is
  required by the usage of anti-unification: we call it to build an existential
  type that would be an upper bound of two input types via abstracting some of
  their subexpressions under existential quantifiers. The existentials quantify
  over \emph{negative} variables, and they must be instantiated in the context
  available at that moment.

  Third, the rule is only applicable if all other rules fail. Notice that it
  could happen even when the input types have matching constructors. For
  example, the generalizer of $[[↑α⁺]]$ and $[[↑β⁺]]$ is $[[γ̂⁻]]$ (with
  mappings $[[γ̂⁻ ↦ ↑α⁺]]$ and $[[γ̂⁻ ↦ ↑β⁺]]$), rather than $[[↑γ̂⁺]]$. This
  way, the algorithm must try to apply the congruent rules first, and only if
  they fail, apply \ruleref{\ottdruleAUAULabel}. As it is not known a priori
  whether the congruent rules will be applicable, the implementation must use 
  backtracking.

@@

This way, the inference algorithm also consists of three mutually recursive
procedures: the positive type inference, the negative type inference, and the
application type inference. As subroutines, the inference algorithm uses
subtyping, constraint merge, and minimal instantiation. The corresponding
dependency is shown in \cref{fig:alg-typing-graph}.

\begin{figure}
  \centering
  \begin{tikzpicture}
    [>={Stealth[scale=2]},node distance=2.2cm,every node/.style={draw,rectangle},every text node part/.style={align=center}]
    % Define nodes
    \node[] (1) {Positive Inference\\$[[Γ; Φ ⊨ v : iP]]$};
    \node[right=1.4cm of 1] (2) {Negative Inference\\$[[Γ; Φ ⊨ c : iN]]$};
    \node[right=1.2cm of 2] (3) {Application Inference\\$[[Γ ; Φ ; Θ1 ⊨ uN ● args ⇒> uM ⫤ Θ2 ; SC]]$};
    \node[below=0.6cm of 2] (7) {Minimal Instantiation\\$[[Γ ⊢ uP SC minby uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[below=0.5cm of 7] (8) {Constraint Singularity\\$[[SC singular with uσ]]$\\(\cref{sec:constraint-singularity})};
    \node[right=0.3cm of 7] (4) {Constraint Merge\\$[[Θ ⊢ SC1 & SC2 = SC3]]$\\(\cref{sec:constraint-merge})};
    \node[left=0.3cm of 7] (6) {Negative Subtyping\\$[[Γ ; Θ ⊨ uN1 ≤ iN2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \node[below=0.9cm of 6] (5) {Positive Subtyping\\$[[Γ ; Θ ⊨ uP1 ≥ iP2 ⫤ SC]]$\\(\cref{sec:subtyping-algorithm})};
    \draw[->] ([yshift=-.2cm]2.west) to ([yshift=-.2cm]1.east);
    \draw[->] ([yshift=+.2cm]1.east) to ([yshift=+.2cm]2.west);
    \draw[->, bend right=20] (3) to (1);
    \draw[->] (2) to (3);
    \draw[->] (2) to (4);
    \draw[->] (2) to (7);
    \draw[->] (3) to (4);
    \draw[->] (7) to (8);
    \draw[->] (1.south west) |- (5);
    \draw[->] (3) |- (5.south east);
    \draw[->] (2) to (6);
  \end{tikzpicture}  
  \caption{Dependency graph of the typing algorithm}
  \label{fig:alg-typing-graph}
\end{figure}

@@

  Rule \ruleref{\ottdruleATCVarLetLabel} infers a type of computational let
  binder. It follows the corresponding declarative rule
  \ruleref{\ottdruleDTCVarLetLabel} but uses algorithmic judgments instead of
  declarative ones. It is worth noting that when calling the subtyping 
  $[[Γ ; · ⊨ uM ≤ ↑iP ⫤ ·]]$, both $[[uM]]$ and $[[iP]]$ are free of algorithmic
  variables: $[[uM]]$ is a type inferred for $[[c]]$, and $[[iP]]$ is given as
  an annotation.

@@

  To ensure the rule precedence, we also require
  the head type to have at least one $[[∀]]$-quantifier, 
  and the list of arguments to be non-empty.

\paragraph{Application of an Arrow Type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the \emph{minimal} constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.

@@

Multiple types $[[iM]]$ can be inferred for a type application:
$[[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ but only one \emph{principal} 
type should be chosen for a variable in an unannotated let binder.
Declaratively, we require the principal type $[[iP]]$ to be
minimal among other all types $[[iP']]$ that can be inferred
for the application $[[Γ ; Φ ⊢ iN ● args ⇒> ↑iP']]$.
Algorithmically, the inference returns an
\emph{algorithmic} type $[[uP]]$ 
together with a set of \emph{necessary and sufficient} 
constraints $[[SC]]$ that any instantiation of $[[uP]]$ must satisfy.
This way, the principal type is the minimal instantiation of $[[uP]]$,
obtained by a substitution satisfying the given constraints $[[SC]]$.

To find this substitution, we use the minimal instantiation
algorithm (\cref{fig:minimal-instantiation}). The judgment `$[[Γ ⊢ uP SC minby uσ]]$'
$[[uσ]]$ instantiates $[[uP]]$ to a subtype of any other 
instantiation of $[[uP]]$ that satisfies $[[SC]]$. 
First, it removes the existential quantifiers by \ruleref{\ottdruleSINGExistsLabel}.
Then it considers two cases. 
\begin{enumerate}
  \item If the type $[[uP]]$ is an algorithmic variable $[[α̂⁺]]$
    restricted by a \emph{subtyping} constraint $[[(α̂⁺ :≥ iQ) ∊ SC]]$,
    its minimal instantiation is the type $[[nf(iQ)]]$, and 
    \ruleref{\ottdruleSINGPUvarLabel} is applied.
  \item Otherwise, the type $[[uP]]$ is either a shifted computation, a
    declarative variable, an unrestricted algorithmic variable, or an
    algorithmic variable restricted by an \emph{equivalence} constraint. In all
    of these cases, the minimal instantiation exists if and only if there is
    only one possible instantiation of $[[uP]]$ that satisfies $[[SC]]$.

    The algorithm finds this instantiation by two premises:
    \begin{enumerate*}
      \item[(i)] checking that all the algorithmic variables of the type are
      restricted by the constraint set; and
      \item[(ii)] building the substitution that satisfies the constraint set on
        these variables (and simultaneously checking that such substitution is
        unique up to equivalence) using the \emph{constraint singularity algorithm}.
    \end{enumerate*}
  \end{enumerate}

@@

\begin{figure}
  \begin{minipage}{0.49\textwidth}
    `$[[SC singular with uσ]]$' means:
    \begin{itemize}[leftmargin=*]
      \item[$+$] for any \emph{positive} constraint entry $[[scE ∊ SC]]$ restricting 
        a variable $[[β̂⁺]]$, there exists $[[iP]]$ such that $[[scE singular with iP]]$
        (as defined in \cref{fig:constraint-entry-singularity}),
        and $[[ [uσ]β̂⁺ = iP ]]$; and
      \item[$-$] the symmetric property holds for all \emph{negative} $[[scE ∊ SC]]$; and
      \item[$\notin$] for any $[[α̂± ∉ dom(SC)]]$, $[[ [uσ]α̂± = α̂± ]]$.
    \end{itemize}
    \caption{Singular Constraint}
    \label{fig:constraint-singularity}
  \end{minipage}
  \begin{minipage}{0.48\textwidth}
    \ottdefnSINGLabeled
    \caption{Singular Constraint Entry}
    \label{fig:constraint-entry-singularity}
  \end{minipage}
\end{figure}

To implement the singularity algorithm, we define a partial function 
`$[[SC singular with uσ]]$', taking a subtyping constraint $[[SC]]$ as an argument and
returning a substitution $[[uσ]]$---the only solution of $[[SC]]$.

The constraint $[[SC]]$ is composed of constraint entries. Therefore, we define
singularity by combining the singularity of each constraint entry
(\cref{fig:constraint-singularity}). In order for $[[SC]]$ to be singular, each
entry must have a unique instantiation, and the resulting substitution $[[uσ]]$
must be a union of these instantiations. In addition, $[[uσ]]$ must act as an
identity on the variables not restricted by $[[SC]]$.

The singularity of constraint \emph{entries} is defined in
\cref{fig:constraint-entry-singularity}. 
\begin{itemize}
  \item The \emph{equivalence} entries are
    always singular, as the only possible type satisfying them is the one given in
    the constraint itself, which is reflected in rules \ruleref{\ottdruleSINGNEqLabel} and
    \ruleref{\ottdruleSINGPEqLabel}. 
  \item The \emph{subtyping} constraints are trickier.
    As will be discussed in \cref{sec:proof-lub-upgrade}, variables (and equivalent
    them $[[∃nas.pa]]$) do not have proper supertypes, and thus, the
    constraints of a form $[[pua :≥ ∃nas.pa]]$ are singular with the only possible
    normalized solution $[[pa]]$---see rule \ruleref{\ottdruleSINGSupVarLabel}. 
  \item However, if the body of the existential type is guarded by a shift $[[∃nas.↓iN]]$, it is
    singular if and only if $[[iN]]$ is equivalent to some $[[αi⁻ ∊ {nas}]]$ bound
    by the quantifier---see rule \ruleref{\ottdruleSINGSupShiftLabel}.
\end{itemize}
